Incremental tactics

### 2012-05-28 doc/papers/LFMTP2012/paper.tex
paper

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for what is commonly refered as ``type checking'',
or explicit ``church-style'' calculi. Take for example dependent type
systems, relying on a certain notion of definitional equality on
types, or type systems with subtype polymorphism: both rely on a rule
of the form
$$
\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}\ \text{,}
$$
for a given notion of $\leq$, which is not syntax-directed. The
implementer then has to convert the declarative system to an
algorithm, and convince himself that both are equivalent. This task,
often neglected, is nonetheless highly non-trivial and non-modular.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to ``forget'' about the declarative system and take only the
result of the algorithm into consideration. But for the sake of the
discussion, let us choose another path: derivations in the declarative
systems, if they cannot always be easily inferred, can be
\emph{verified} if provided by an oracle. Even if the oracle is wrong,
we can tell cheaply by verifying the witness typing derivation it
provides. Let us then imagine a type-checker to be the pair of 1/ an
untrusted procedure $\finfer M$ for inferring a typing derivations out
of programs $M$; it doesn't have to be complete, or even sound, and 2/
a trusted and fast kernel $\fcheck{\md}$ for verifying these
derivations in a given set of inference rules $\Sigma$. This is a
certifying scheme: if $\fcheck{\finfer M} = \cst{true}$ then there
really \emph{is} a derivation $\md$ for $M$. Of course the converse is
not necessarily true: function $\finfer{}\!\!$ could fail, or have
bugs (return an incorrect derivation). Explicitly generating the
derivation in the declarative system has the following advantages over
the first approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\item for optimization purpose, compilers often need to propagate type
  information in the chain of transformation; the typing derivation
  produced by $\finfer{}$ being explicit, it could constitute the data
  structure being exploited.
\end{itemize}
\ldots\ but there is more:

@@

Compilers are generally monolithic programs, taking source code and
producing machine code in a series of transformations, one of the
first being type checking. But as type systems become richer, it
becomes increasingly difficult for the programmer to write a correct
program in one try, compile it and run it: writing a program longer
than a few dozen of lines becomes a tight and non-linear interaction
between the human and the type-checker involving constant experiments,
fixes, etc. If this interaction is possible for short programs and
fast checkers, it becomes increasingly hindered by the latency of
complete rechecks: even if the modifications made to the program
between two interactions are small, the whole program will be
rechecked. This is especially true for languages embedding formal
verification aspects, like proof assistants: there, tight interaction
with a type or proof checker is unavoidable due to the difficulty of
writing correct proofs without guidance from the system, and the time
taken by the tool to infer trivial parts of the proofs (proof
search). The constant modification of the source makes it necessary to
observe \emph{incrementally} the effect of a small change on the whole
edifice: we need to take advantage of the results of previous checks
in order to build the result for a modified version of the program.

%% State your contributions (Bulleted list of contributions)

@@

A student (the first author) is learning about System \sysname T, and
wants to write a program computing $(2+2)\times 3$. He knows that
sometimes those $\cst{rec}$ arguments are quite hard to get correct,
but System \sysname T is a typed language; he wants to take advantage
of this to build his program incrementally using the teacher's (the
first author too) provided incremental verifier. As a first attempt,
he wants to verify $P_1$:
$$
\recb{\s\z}{\s\z}{\lam x\lam y \s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
  x}$, and he could reuse the pieces of derivation just built. Let us
call these $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
$\get\md$ from the derivation to the program it types, what needs
really to be sent to the system is:
\begin{align*}
  &\finfer{(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} {\get{\md_4}}}
    \app{\var{add}}\app{\get{\md_2}}{\get{\md_2}}
  )}
\ \text{,}
\end{align*}
that is, only the subterms of the program that changed. Derivations
$\md_2$ and $\md_4$ used to live in the empty context, we are then
free to use it in the new context where
$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
the \emph{weakening} property. Taking advantage of $\md_2$ and $\md_4$
the system succeeds with a new derivation made out of these, without
having to regenerate them because it knows that $\finfer{\get\md} =
\md$. Let us call $\md_5$ the subderivation generated for $\tlam x
{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
{\get{\md_4}}$, $\md_6$ for the arguments
$\app{\get{\md_2}}{\get{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\get{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\get{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\get{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\get{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so just use it.

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\finfer{}(
\letb{add}{\get{\md_5}}
\get{({\gsubst{\md_9} {\msubst {H} {(\get{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\get{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one in the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\get{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\&\quad
  {\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}}}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

Although this small example may seem anecdotal, the interest of a
mechanism for reusing results of derivation generation becomes clear
on very large programs: then the cost of rechecking well-typing of a
small modification amounts to recheking the path in the term from the
root to the changed part, and the changed part itself.

A type-checker is usually viewed as an algorithm returning a binary
answer (ok/error), specified by a set of inference rules. Seeing
type-checking as issueing a completely explicit typing derivation
enables to think of the relationship between the input and the
output. Feeding part of the output $\md_i$ back to the input $P_i$ is
a well-known technique for handling changes in the input: combined
with a way to automatically recognize already-computed input, it is
\emph{memoization}, a general-purpose technique for incremental
computation that has been extensively studied and combined with other
techniques \cite{pugh1989incremental, acar2003selective}. However, it
is best understood in the case of purely first-order data, and to our
best knowledge remains unexplored for higher-order structures
containing binders, like programs and proofs. In this case,

To simplify the problem,
we separate the task in:
\begin{itemize}
\item from a modified program, generate the \emph{diff} against the
  previous version, that is, observing and linking to unchanged
  derivations.
\item given a \emph{diff}, construct a new derivation and verify it.
\end{itemize}
The latter is the subject of this paper.

@@

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. The requirements are:
\begin{itemize}
\item conciseness of the certificates, since they are usually stored
  or transfered remotely; in particular,
\item a well-understood, short and independent verification algorithm;
  this is known as the \emph{de Bruijn principle}.
\item universality with respect to the particular problem: having a
  certificate language for each domain of application would just shift
  the problem of trust; having one universally accepted certificate
  language that can be tailored to many problems puts a much stronger
  trust on the unique verifier.
\end{itemize}

The Edimburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is a well-accepted solution to this question. It is a typed
lambda-calculus and can be viewed as a language for
\emph{representing} terms, derivations and statements of $\Pi_0^1$
meta-theorems (but not their proofs) in a given inference system
supporting hypothetical judgements \cite{pfenning2001logical}. For
these, it provides built-in support for encoding languages with
binders or hypothesis (a feature called $\lambda$-tree syntax or HOAS
\cite{pfenning1988higher}). The user gives a \emph{signature} defining
abstract syntax and inference rules of his \emph{object language}, and
gets an algorithm to verify terms and derivations expressed in the
\LF\ syntax.

@@

We overload \emph{substitution} as a partial operation on objects, given
a parallel substitution $\sigma$. Since the syntax of object is not
stable by (textual) substitution, it is defined \emph{hereditarily}: it
is mutually recursive with a \emph{cut} function $\gcut M S$ which can
in turn trigger a chain of substitutions, leading to a canonical objects
or an error.
\begin{align*}
  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
    \sigma)}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \app {\var x} {\gsubst S \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \gcut M {\gsubst S \sigma} &
  \text{if $\sigma(\var x) = M$}
  \\
  \gsubst {\smeta X {\sigma}} {\sigma'} &=
  \smeta X {\gcomp\sigma{\sigma'}}
  \\
  \gsubst\spinenil\sigma &=
  \spinenil \\
  \gsubst{\spinecons M S}\sigma &=
  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
  \\
  \gcut {\lam x M} {\spinecons N S} &=
  \gcut {\gsubst M {\msubst x N}} S
  \\
  % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
  % \gcut {\smeta X\sigma} S &=
  % \smeta X{\gcomp\sigma\sigma'}
  % \\
  \gcut {\app H S} \spinenil &=
  \app H S
  \\
  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
  \\
  \gcomp \sigma \msubstnil &=
  \sigma
  \\
  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
  \\
  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
  \\
  \gsubst {\msubstnil}{\sigma} &= \msubstnil
\end{align*}

@@

Note that this notion of substitution differs from hereditary
substitution found in \eg\ \cite{hl07mechanizing} as it presupposes
$\eta$-long normal forms: once a cut starts, there must be exactly as
many formal and actual arguments. A non-empty spine cut against
another non-empty spine $\gcut {\app H S} {S'}$ blocks the
substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
implicit $\eta$-expansion); conversely, exhausting actual but not
formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
substitution too. For example, the reduction $\gcut {\lam x \app {\var
    x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck because the
second $\var x$ in the left part is substituted by a function yet is
not itself a function.

@@

The algorithm is presented in Fig. \ref{fig:obj-typing} as a
syntax-directed inference system. All judgements are of the form $\gjo
U V W$ where $U$ and $V$ are inputs, $V$ is the induction hypothesis
and $W$ the output. They are parameterized by a constant and implicit
signature $\Sigma$, and all premisses are ordered in the algorithmical
order of computation.

>>>>

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for what is commonly referred as ``type checking'',
or explicit ``church-style'' calculi. Take for example dependent type
systems, relying on a certain notion of definitional equality on
types, or type systems with subtype polymorphism: both rely on a rule
of the form
$$
\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}\ \text{,}
$$
for a given notion of $\leq$, which is not syntax-directed. The
implementer then has to convert the declarative system to an
algorithm, and convince himself that both are equivalent. This task,
often neglected, is nonetheless highly non-trivial and non-modular.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to ``forget'' about the declarative system and take only the
result of the algorithm into consideration. But for the sake of the
discussion, let us choose another path: derivations in the declarative
systems, if they cannot always be easily inferred, can be
\emph{verified} if provided by an oracle. Even if the oracle is wrong,
we can tell cheaply by verifying the witness typing derivation it
provides. Let us then imagine a type-checker to be the pair of 1/ an
untrusted procedure $\finfer M$ to infer a typing derivations out of
programs $M$; it doesn't have to be complete, or even sound, and 2/ a
trusted and fast kernel $\fcheck{\md}$ to verify these derivations in
a given set of inference rules $\Sigma$. This is a \emph{certifying}
scheme: if $\fcheck{\finfer M} = \cst{true}$ then there really
\emph{is} a derivation $\md$ for $M$. Of course the converse is not
necessarily true: function $\finfer{}\!\!$ could fail, or have bugs
(return an incorrect derivation). Explicitly generating the derivation
in the declarative system has the following advantages over the first
approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\item for optimization purpose, compilers often need to propagate type
  information in the chain of transformation; the typing derivation
  produced by $\finfer{}$ being explicit, it could constitute the data
  structure being exploited.
\end{itemize}
\ldots\ but there is more:

@@

As type systems become richer, it becomes increasingly difficult for
the programmer to write a correct program in one try, compile it and
run it: writing a program longer than a few dozen of lines becomes a
tight and non-linear interaction between the human and the
type-checker involving constant experiments, fixes, etc. If this
interaction is possible for short programs and fast checkers, it
becomes increasingly hindered by the latency of complete rechecks:
even if the modifications made to the program between two interactions
are small, the whole program is usually rechecked. This is especially
true for languages embedding formal verification aspects, like proof
assistants: there, tight interaction with a type or proof checker is
unavoidable due to the difficulty of writing correct proofs without
guidance from the system, and the time taken by many tools to infer
easy parts of the proofs (proof search). The constant modification of
the source makes it necessary to observe \emph{incrementally} the
effect of a small change on the whole edifice: we need to take
advantage of the results of previous checks in order to build the
result for a modified version of the program.

Several methods have been devised for turning a batch procedure into
an incremental one \cite{pugh1989incremental,
  acar2003selective}. Unfortunately, they all suffer from their
genericity on the particular problem of type checking:
\begin{itemize}
\item first, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  ground derivation of $\vdash 2+2 : \cst{nat}$ is still valid when
  put under a binder $\vdash\lam x 2 + 2 - x:\cst{nat}\to\cst{nat}$
  and should not be regenerated;
\item for very large programs, we might want to feed the type checker
  only with a small \emph{delta} on the previous version and not the
  whole program, thus saving the time to recognize already checked
  parts;
\item finally, there is no way to \emph{a posteriori} validate the
  result of the incremental process: the user has to trust not only
  the type-checking algorithm but also its transformation into an
  incremental one.
\end{itemize}
It just so happens that, if we consider as above a type-checker as a
derivation-generating procedure, and storing these derivations between
calls to the procedure, we could recognize already-built
subderivations for subterms common to a checked and a new version to
check of the program, and graft these pieces of derivation to form a
new, large one for the whole modified program, saving the regeneration
of all equal subterms' derivations in compatible contexts. This way we
maintain the verifiable witness of well-typing for the program, and
take advantage of the quotients made on higher-order syntax.

We can thus add to the previous list that explicitly generating typing
derivations makes possible a form of safe incrementality by sharing
common subderivations.

In this paper, we present a framework to incrementally generate typing
certificates represented in the \LF\ logical framework, and implemented
as an \sysname{ML} library. We specifically propose:
\begin{itemize}
\item a simple language for representing programs and proof
  deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
\item an untyped, unsafe calculus to write derivation-generating type
  checkers, based on a notion of function inverse %TODO le nom?
  to refer an already-computed value;
\item a generic algorithm to evaluate these type-checkers on program
  deltas and verify on-the-fly the generated certificates
  incrementally against a \emph{repository} of already constructed and
  checked derivations.
\end{itemize}

@@

A student is learning about System \sysname T, and wants to write a
program computing $(2+2)\times 3$. He knows that sometimes those
$\cst{rec}$ arguments are quite hard to get correct, but System
\sysname T is a typed language; he wants to take advantage of this to
build his program incrementally using the teacher's provided
incremental verifier. As a first attempt, he wants to verify $P_1$:
$$
\recb{\s\z}{\s\z}{\lam x\lam y \s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
  x}$, and he could reuse the pieces of derivation just built. Let us
call these $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
$\fget\md$ from the derivation to the program it types, what needs
really to be sent to the system is:
\begin{align*}
  &\finfer{(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} {\fget{\md_4}}}
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )}
\ \text{,}
\end{align*}
that is, only the subterms of the program that changed. Derivations
$\md_2$ and $\md_4$ used to live in the empty context, we are then
free to use it in the new context where
$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
the \emph{weakening} property. Taking advantage of $\md_2$ and $\md_4$
the system succeeds with a new derivation made out of these, without
having to regenerate them because it knows that $\finfer{\fget\md} =
\md$. Let us call $\md_5$ the subderivation generated for $\tlam x
{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
{\fget{\md_4}}$, $\md_6$ for the arguments
$\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\fget{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\fget{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\fget{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so just use it.

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\finfer{}(
\letb{add}{\fget{\md_5}}
\fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one in the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\fget{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\&\quad
  {\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}}}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

We choose here to represent deltas as usual terms with variables $\var
x, \var y$ and special, \emph{meta}-variables referring to
already-computed results $\md_i$. Commands sent to the system consists
of terms, possibly containing special typed operators having a
computational content: $\finfer{\cdot}$ takes a term $M$ and produces
the (dependent) pair of a type $A$ and a derivation of the judgment
$\vdash M:A$ in the current context; $\fget{\cdot}$ does the exact
converse: it takes a derivation $\vdash M:A$ and projects it back as
the term $M$. The successful result of a command execution is a
derivation where each subderivation has been given a unique
metavariable name for later reuse (a \emph{repository}). As
subderivations can have free variables, we close them by a local
environment that must be instantiated by a \emph{substitution}.

Although this small example may seem anecdotal, the interest of such a
mechanism becomes clear on very large programs: then the cost of
rechecking well-typing of a small modification amounts to rechecking
the path in the term from the root to the changed part, and the
changed part itself.

%TODO figure?

Seeing type-checking as issuing a completely explicit typing
derivation enables thus to think of the relationship between the input
and the output. Feeding subterms of the output $\md_i$ back into
subterms of the input $P_i$ is a well-known technique for handling
changes in the input: combined with a way to automatically recognize
already-computed input, it is \emph{memoization}, a general-purpose
technique for incremental computation that has been extensively
studied and combined with other techniques \cite{pugh1989incremental,
  acar2003selective}. However, it is best understood in the case of
purely first-order data, and to our best knowledge remains unexplored
for higher-order structures containing binders, like programs and
proofs.

A memoized type-checker stores in a table bindings from its input (the
program) to its output (the derivation), and returns automatically the
stored output if the current input matches a binding in the table. It
thus performs two operations at once: output reuse and recognition of
already-seen input. Since       %TODO

To simplify the problem,
we separate the task in:
\begin{itemize}
\item from a modified program, generate the \emph{delta} against the
  previous version, that is, observing and linking to unchanged
  derivations.
\item given a \emph{delta}, construct a new derivation and verify it.
\end{itemize}
The latter is the subject of this paper.

@@

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. The requirements are:
\begin{itemize}
\item conciseness of the certificates, since they are usually stored
  or transferred remotely; in particular,
\item a well-understood, short and independent verification algorithm;
  this is known as the \emph{de Bruijn principle}.
\item universality with respect to the particular problem: having a
  certificate language for each domain of application would just shift
  the problem of trust; having one universally accepted certificate
  language that can be tailored to many problems puts a much stronger
  trust on the unique verifier.
\end{itemize}

The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is a well-accepted solution to this question. It is a typed
lambda-calculus and can be viewed as a language for
\emph{representing} terms, derivations and statements of $\Pi_0^1$
meta-theorems (but not their proofs) in a given inference system
supporting hypothetical judgments \cite{pfenning2001logical}. For
these, it provides built-in support for encoding languages with
binders or hypothesis (a feature called $\lambda$-tree syntax or HOAS
\cite{pfenning1988higher}). The user gives a \emph{signature} defining
abstract syntax and inference rules of his \emph{object language}, and
gets an algorithm to verify terms and derivations expressed in the
\LF\ syntax.

@@

We overload \emph{substitution} as a partial operation on objects,
given a parallel substitution $\sigma$. Since the syntax of object is
not stable by (textual) substitution, it is defined as
\emph{hereditary}: it is mutually recursive with a \emph{cut} function
$\gcut M S$ which can in turn trigger a chain of substitutions,
leading to a canonical objects or an error.
\begin{align*}
  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
    \sigma)}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \app {\var x} {\gsubst S \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \gcut M {\gsubst S \sigma} &
  \text{if $\sigma(\var x) = M$}
  \\
  \gsubst {\smeta X {\sigma}} {\sigma'} &=
  \smeta X {\gcomp\sigma{\sigma'}}
  \\
  \gsubst\spinenil\sigma &=
  \spinenil \\
  \gsubst{\spinecons M S}\sigma &=
  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
  \\
  \gcut {\lam x M} {\spinecons N S} &=
  \gcut {\gsubst M {\msubst x N}} S
  \\
  % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
  % \gcut {\smeta X\sigma} S &=
  % \smeta X{\gcomp\sigma\sigma'}
  % \\
  \gcut {\app H S} \spinenil &=
  \app H S
  \\
  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
  \\
  \gcomp \sigma \msubstnil &=
  \sigma
  \\
  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
  \\
  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
  \\
  \gsubst {\msubstnil}{\sigma} &= \msubstnil
\end{align*}

@@

Note that this notion of substitution differs from hereditary
substitution found in \eg\ \cite{hl07mechanizing} as it presupposes
$\eta$-long normal forms: once a \emph{cut} starts, there must be
exactly as many formal and actual arguments. A non-empty spine cut
against another non-empty spine $\gcut {\app H S} {S'}$ blocks the
substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
implicit $\eta$-expansion); conversely, exhausting actual but not
formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
substitution too. For example, the reduction $\gcut {\lam x \app {\var
    x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck because the
second $\var x$ in the left part is substituted by a function yet is
not itself a function.

@@

The algorithm is presented in Fig. \ref{fig:obj-typing} as a
syntax-directed inference system. All judgments are of the form $\gjo
U V W$ where $U$ and $V$ are inputs, $V$ is the induction hypothesis
and $W$ the output. They are parameterized by a constant and implicit
signature $\Sigma$, and all premises are ordered in the algorithmic
order of computation.
